---
title: "Reproducible Research"
date: "Coursera Course Assignment"
author: "Nicholas Marino"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning=  FALSE)
```

# Loading and preprocessing the data

Loading the packages to analyze the data.

```{r}
# loading packages
library(tidyverse)
```

Loading the dataset

```{r}
# loading the data
df <- read_csv(file = 'activity.csv')
df
```

Skimming the data.

```{r}
skimr::skim(data = df)
```

# What is mean total number of steps taken per day?

1. Calculate the total number of steps taken per day

```{r}
df %>% 
  group_by(date) %>% 
  summarise(total_steps = sum(steps, na.rm = TRUE))
```

2. Make a histogram of the total number of steps taken each day

```{r fig.align='center', fig.width=8, fig.height=5}
df %>% 
  group_by(date) %>% 
  summarise(total_steps = sum(steps, na.rm = TRUE)) %>% 
  pull(total_steps) %>% 
  hist(x = ., main = 'Total number of step taken each day', xlab = 'Total number of steps')
```

3. Calculate and report the mean and median of the total number of steps taken per day

```{r fig.align='center', fig.width=8, fig.height=5}
df %>% 
  group_by(date) %>% 
  summarise(
    mean_steps   = mean(steps, na.rm = TRUE),
    median_steps = median(steps, na.rm = TRUE)
  ) %>% 
  pivot_longer(cols = -date, names_to = 'variable', values_to = 'values') %>% 
  ggplot(mapping = aes(x = date, y = values, color = variable)) +
  facet_wrap(~ variable, ncol = 2, scales = 'free_y') +
  geom_line(size = 1, show.legend = FALSE) +
  labs(x = 'Steps', y = 'Date') +
  theme_minimal()
```

# What is the average daily activity pattern?

1. Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r fig.align='center', fig.width=8, fig.height=5}
df %>% 
  group_by(interval) %>% 
  summarise(
    mean_steps   = mean(steps, na.rm = TRUE)
  ) %>% 
  ggplot(mapping = aes(x = interval, y = mean_steps)) +
  geom_line(size = 1) +
  labs(x = 'Interval', y = 'Steps') +
  theme_minimal()
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
df %>% 
  group_by(interval) %>% 
  summarise(
    mean_steps   = mean(steps, na.rm = TRUE)
  ) %>% 
  slice_max(n = 1, order_by = mean_steps)
```

# Imputing missing values

1. Calculate and report the total number of missing values in the dataset

```{r}
df %>% 
  is.na() %>% 
  colSums
```

2. Devise a strategy for filling in all of the missing values in the dataset

```{r}
df %>% 
  group_by(interval) %>% 
  mutate(steps = ifelse(test = is.na(steps), 
                        yes = round(x = mean(steps, na.rm = TRUE), digits = 0), no = steps))
```

3. Create a new dataset that is equal to the original dataset but with the missing data filled in

```{r}
df_imputed <- df %>% 
  group_by(interval) %>% 
  mutate(steps = ifelse(test = is.na(steps), 
                        yes = round(x = mean(steps, na.rm = TRUE), digits = 0), no = steps)) %>% 
  ungroup
df_imputed
```

4. Make a histogram of the total number of steps taken each day and calculate and report the mean and median total number of steps taken per day.

```{r fig.align='center', fig.width=8, fig.height=5}
df_imputed %>% 
  group_by(date) %>% 
  summarise(total_steps = sum(steps, na.rm = TRUE)) %>% 
  pull(total_steps) %>% 
  hist(x = ., main = 'Total number of step taken each day (imputed)', xlab = 'Total number of steps')
```

a. Do these values differ from the estimates from the first part of the assignment?

Yes.

b. What is the impact of imputing missing data on the estimates of the total daily number of steps?

Data became more normally distributed, with much more observations towards 10K to 15K steps.

# Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day

```{r}
df <- df %>% 
  mutate(
    wweek = weekdays(date),
    wweek = case_when(wweek %in% c('Saturday', 'Sunday') ~ 'weekend',
                      TRUE ~ 'weekday')
  )
df
```

2. Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r fig.align='center', fig.width=8, fig.height=5}
df %>% 
  group_by(interval, wweek) %>% 
  summarise(
    mean_steps = mean(steps, na.rm = TRUE),
    .groups = 'drop'
  ) %>% 
  ggplot(mapping = aes(x = interval, y = mean_steps, color = wweek)) +
  facet_wrap(~ wweek, ncol = 2, scales = 'free') +
  geom_line(size = 1, show.legend = FALSE) +
  labs(x = 'Interval', y = 'Steps') +
  theme_minimal()
```
